\documentclass[entropy,article,submit,pdftex,moreauthors]{Definitions/mdpi} 

%=================================================================
% MDPI internal commands 
\firstpage{1} 
\makeatletter 
\setcounter{page}{\@firstpage} 
\makeatother
\pubvolume{1}
\issuenum{1}
\articlenumber{0}
\pubyear{2025}
\copyrightyear{2025}
\datereceived{ } 
\daterevised{ }
\dateaccepted{ } 
\datepublished{ } 
\hreflink{https://doi.org/10.5281/zenodo.17810561}

\setlength{\headheight}{25pt}

%=================================================================
% Full title of the paper (Capitalized)
\Title{A Scalar Diagnostic for Empirical Score Alignment on Fisher Manifolds}

\TitleCitation{Scalar Diagnostic for Empirical Score Alignment}

\newcommand{\orcidauthorA}{0009-0008-5550-9716} 

\Author{Isaid Cornejo $^{1}$\orcidA{}}

\AuthorNames{Isaid Cornejo}

\AuthorCitation{Cornejo, I.}

\address{%
$^{1}$ \quad Information Physics Institute (IPI); isaidcor@gmail.com
}

\corres{Correspondence: isaidcor@gmail.com; Tel.: +52-56-4272-4979}

% Abstract
\abstract{
    The Fisher--Rao metric provides the canonical geometric structure on statistical
    manifolds, yet empirical data often induce strong deviations from this geometry,
    manifesting through heavy-tailed spectra, dominant outlier modes, and effective
    dimensionality collapse in modern statistical models and neural networks. Existing
    metrics for quantifying these effects lack reparametriza\-tion invariance and do not
    directly compare empirical sensitivity to the intrinsic geometric baseline. We
    introduce a coordinate-free scalar diagnostic that measures the empirical deformation
    of Fisher geometry by contracting the empirical score covariance with the inverse
    Fisher metric, together with a rectified amplitude that isolates the excess-alignment
    sector. The associated alignment operator is shown to be diagonalizable with real
    non-negative eigenvalues, and the scalar diagnostic aggregates deviations through an
    invariant spectral identity that distinguishes empirical reinforcement from suppression.
    We validate the diagnostic across exponential families, Gaussian mixture models, and a
    trained neural network on MNIST, demonstrating that it captures equilibrium,
    suppression, and strong empirical reinforcement in a unified scalar signature.
    Reproducible code and figures are provided online.
}

\keyword{Fisher information; information geometry; empirical score covariance; alignment operator; spectral diagnostics; statistical manifolds; eigenvalue analysis; neural networks; high-dimensional statistics}
\MSC{62B10; 62H12; 68T07}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}

\input{sections/2_introduction}
\input{sections/3_background}
\input{sections/4_diagnostic}
\input{sections/5_spectral}
\input{sections/6_regimes}
\input{sections/7_experiments}
\input{sections/8_discussion}
\input{sections/9_conclusion}

\section*{Summary Table of Spectral Regimes}
\input{tables/summary_table}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\authorcontributions{
Conceptualization, I.C.; methodology, I.C.; software, I.C.; validation, I.C.; formal analysis, I.C.; investigation, I.C.; resources, I.C.; data curation, I.C.; visualization, I.C.; writing---original draft preparation, I.C.; writing---review and editing, I.C. The author has read and agreed to the published version of the manuscript.
}

\funding{
This research received no external funding.
}

\institutionalreview{
Not applicable.
}

\informedconsent{
Not applicable.
}

\dataavailability{
All code and reproducibility materials supporting this study are openly available in the GitHub repository: \url{https://github.com/isaidcornejo/coherence-field}.
}

\acknowledgments{
Not applicable.
}

\conflictsofinterest{
The author declares no conflict of interest.
}

\begin{adjustwidth}{-\extralength}{0cm}

\reftitle{References}
\bibliography{../references}


\PublishersNote{}
\end{adjustwidth}

\end{document}